---
title: "Konsistensmidler information request"
author: "Trine Husøy"
date: "2024-11-18"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
  pdf_document:
    toc: true
fontsize: 10pt
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, warning = FALSE, error = FALSE, message = FALSE, fig.width = 6)
```

# Loading packages

To get the ggsanky package I needed to install a development version from github, which needed devtools.
```{r}

#install.packages("patchwork")

library(ggplot2)
library(openxlsx)
library(tidyverse)
library(ggsankey)
library(patchwork)
```

# Get work directory and organise results

```{r}

HOME <- "C:/Users/TRHU/Documents/R_projects/Fig_oppdate_konsistens"
setwd(HOME)

```

Create a folder with current date in the Result folder
```{r}

newday <- file.path('C:/Users/TRHU/Documents/R_projects/Fig_oppdate_konsistens/Results', Sys.Date())
dir.create(newday)

```


# Read in data
```{r}
request <- read_csv2("./Data/Konsistensmiddler_information_request.csv")
animal <- read_csv2("./Data/Data_animal_030325.csv")
human <- read.csv2("./Data/Data_human_030325.csv")


```

# Data cleaning

## Sankey plot

Prepare data file for the Sankey plot by changing the name of the content in the coloumns and the heading of the coloumns.

```{r}


request <- as_tibble(request)

request <- request %>% 
  rename(`No e-mail` = `No email for corresponding author in publication/ Email for corresponding author no longer working`)

request$`No e-mail`[is.na(request$`No e-mail`)] <- "y" # Need to change NA's to y, because if_else don't work with NA's
request$`No answer`[is.na(request$`No answer`)] <- "y"
request$Answer[is.na(request$Answer)] <- "y"

request <- request %>% 
  mutate(Request = if_else(`No e-mail`== "x", "No e-mail (N = 119)",
                           if_else(`No answer`== "x", "No answer (N = 36)", "Answer (N = 31)")))

request <- request %>% 
  mutate(Status =if_else(`Included/ excluded` == "Include", "Include (N = 5)", "Exclude (N = 26)"))


request$Source <- "Publications (N=186)"



```

The next code organise the data to make it suitable as input for ggsankey package, to make the Sankey plot. 
```{r}

s1 <- request$Source

s2 <- request$Request

s3 <- request$Status




d <- data.frame(cbind(s1,s2,s3))
names(d) <- c('Source', 
              'Request', 
              'Status')


df <- d%>%
  make_long(Source, 
           Request, 
           Status
           )
```

## Animal and human data


### Cleaning the animal data with regard to the chemical tested

Delete rows and columns with only NAs
```{r}
# animal data
animal <- animal[,1:11] #select the columns with data
animal <- animal[1:42,] # select the rows

```



```{r}
# Here I search for the different substances in the text of the coloumn `Substance, dose and exposure frequency`, and make a new coloumn
# with the heading of the substance and with "yes" or "no" if the study testing the substance.

animal_substance <- animal %>%  select(1:3, 10) # Select the coloums I will use to get which chemical is tested

animal_substance <- na.omit(animal_substance) # delete the rows with NAs

animal_substance <- animal_substance %>% 
  mutate(`Carrageenan (E 407)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Carrageenan"), "yes", "No"))

animal_substance <- animal_substance %>% 
  mutate(`Carboxymethyl cellulose (E 466)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Carboxymethyl cellulose"), "yes", "No"))

animal_substance <- animal_substance %>% 
  mutate(`Agar (E 406)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Agar"), "yes", "No"))
  
animal_substance <- animal_substance %>% 
  mutate(`Guar gum (E 412)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Guar gum"), "yes", "No"))

animal_substance <- animal_substance %>% 
  mutate(`Sodium alginate (E 401)`= if_else(str_detect(`Substance, dose and exposure frequency`, "alginate"), "yes", "No")) 

animal_substance <- animal_substance %>% 
  mutate(`Eucheuma seaweed (E 407a)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Eucheuma seaweed"), "yes", "No"))

animal_substance <- animal_substance %>% 
  mutate(`Gellan gum (E 418)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Gellan gum"), "yes", "No"))

animal_substance <- animal_substance %>% 
  mutate(`Xanthan gum (E 415)`= if_else(str_detect(`Substance, dose and exposure frequency`, "Xanthan gum"), "yes", "No"))


animal_substance <- dplyr::select(animal_substance, c(!`Substance, dose and exposure frequency`)) # Delete the messy coloumn `Substance, dose and exposure frequency`
animal_substance <- dplyr::select(animal_substance, c(!`Primary study`))

animal_substance_long <- animal_substance %>% pivot_longer(
  cols = "Carrageenan (E 407)":"Xanthan gum (E 415)",
  names_to = "Substance",
  values_to = "Included"
)

animal_substance_long<- animal_substance_long %>% 
  filter(Included == "yes")

animal_substance_long$species <- "animal"


animal_substance_long <- animal_substance_long %>% 
  mutate(Year_cat = if_else(Year < 1993, 
                            "<1993",
                            if_else(Year >= 1994 & Year <= 2003, 
                                    "1994-2003",
                                    if_else(Year >= 2004 & Year <= 2013, 
                                            "2004-2013",
                                            if_else(Year >= 2014 & Year <= 2024, 
                                                    "2014-2024", 
                                                    "False")))))




```

### Cleaning the human data with regard to the chemical tested

```{r}
human <- human[1:5,] # select rows


```


```{r}

human_substance <- human %>% dplyr::select(c("Reference", "Year", "Substance"))

human_substance <- human_substance %>% 
  mutate(`Carrageenan (E 407)`= if_else(str_detect(Substance, "Carrageenan"), "yes", "No"))

human_substance <- human_substance %>% 
  mutate(`Carboxymethyl cellulose (E 466)`= if_else(str_detect(Substance, "Carboxymethyl cellulose"), "yes", "No"))

human_substance <- human_substance %>% 
  mutate(`Agar (E 406)`= if_else(str_detect(Substance, "Agar"), "yes", "No"))
  
human_substance <- human_substance %>% 
  mutate(`Guar gum (E 412)`= if_else(str_detect(Substance, "Guar gum"), "yes", "No"))

human_substance <- human_substance %>% 
  mutate(`Sodium alginate (E 401)`= if_else(str_detect(Substance, "alginate"), "yes", "No")) 

human_substance <- human_substance %>% 
  mutate(`Eucheuma seaweed (E 407a)`= if_else(str_detect(Substance, "Eucheuma seaweed"), "yes", "No"))

human_substance <- human_substance %>% 
  mutate(`Gellan gum (E 418)`= if_else(str_detect(Substance, "Gellan gum"), "yes", "No"))

human_substance <- human_substance %>% 
  mutate(`Xanthan gum (E 415)`= if_else(str_detect(Substance, "Xanthan gum"), "yes", "No"))

human_substance <- dplyr::select(human_substance, c(!Substance))

human_substance_long <- human_substance %>% pivot_longer(
  cols = "Carrageenan (E 407)":"Xanthan gum (E 415)",
  names_to = "Substance",
  values_to = "Included"
)

human_substance_long<- human_substance_long %>% 
  filter(Included == "yes")

human_substance_long$species <- "human"


human_substance_long <- human_substance_long %>% 
  mutate(Year_cat = if_else(Year < 1993, 
                            "<1993",
                            if_else(Year >= 1994 & Year <= 2003, 
                                    "1994-2003",
                                    if_else(Year >= 2004 & Year <= 2013, 
                                            "2004-2013",
                                            if_else(Year >= 2014 & Year <= 2024, 
                                                    "2014-2024", 
                                                    "False")))))
```



### Combine the data sets

```{r}
# Combine the data file for animals and humans

animal_human_substance_year <- rbind(animal_substance_long, human_substance_long)
```

Save the Excel file

```{r}
write.xlsx(animal_human_substance_year,
           file = file.path(newday,"animal_human_substance_year.xlsx"),
           colNames = TRUE, borders = "rows"
)
```


### Cleaning the animal data with regard to the health outcome

```{r}
animal_outcome <- animal

animal_outcome <- animal_outcome %>% # delete row nr 5
  slice(-5)

animal_outcome <- animal_outcome %>% 
  select("Reference", "Year", "Outcome as reported") # Select the coloumns of interest

animal_outcome <- animal_outcome %>% 
  slice(-19)


animal_outcome <- animal_outcome %>% # Fill in the NA cells with the content of the cells above
  fill(Reference, .direction = "down")
animal_outcome <- animal_outcome %>% 
  fill(Year, .direction = "down")


```


The standardised outcome categories were misspelled, not harmonised etc, and therefore the outcome were categorised to coloumns with harmonised outcome names. These can be used to make a long table and the names will be simmilar and can be counted
```{r}
animal_outcome <- animal_outcome %>% 
  mutate(`Changes in gut microbiota composition or number of bacteria`= if_else(str_detect(`Outcome as reported`, "Changes in gut microbiota"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Enzymatic activity (digestive, microbial, cell proliferation, inflammation)`= if_else(str_detect(`Outcome as reported`, "Enzymatic activity"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Faecal or caecal content weight, colour, consistency, viscosity`= if_else(str_detect(`Outcome as reported`, "Faecal or caecal content"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Intestinal permeability (markers in serum)`= if_else(str_detect(`Outcome as reported`, "Intestinal permeability"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Intestinal utilisation and fermentation of nutrients`= if_else(str_detect(`Outcome as reported`, "Intestinal utilisation"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Macroscopic changes (stomach, small intestine, colon)`= if_else(str_detect(`Outcome as reported`, "Macroscopic changes"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Microscopic changes (digestive tract) including inflammation`= if_else(str_detect(`Outcome as reported`, "Microscopic changes"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Mucosal weight, content (colon)`= if_else(str_detect(`Outcome as reported`, "Mucosal weight"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Presence of blood or mucus in faeces`= if_else(str_detect(`Outcome as reported`, "Presence of blood"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Tumour promotion in cancer-induced mice (genetic or chemical)`= if_else(str_detect(`Outcome as reported`, "Tumour"), "yes", "No"))

animal_outcome <- animal_outcome %>% 
  mutate(`Weight, length of stomach, small intestine, large intestine`= if_else(str_detect(`Outcome as reported`, "Weight"), "yes", "No"))

# The original "outcome" coloumn is deleted and a new one is made based on the headings of the new coloumns by making a long table

animal_outcome <- dplyr::select(animal_outcome, c(!`Outcome as reported`))

animal_outcome <- animal_outcome %>% pivot_longer(
  cols = "Changes in gut microbiota composition or number of bacteria":"Weight, length of stomach, small intestine, large intestine",
  names_to = "outcome",
  values_to = "Included"
)

animal_outcome <- animal_outcome %>% 
  filter(Included == "yes")

animal_outcome_wide <- animal_outcome %>% pivot_wider(
  names_from = "outcome",
  values_from = "Included"
)


```


Merge the animal tables with substance_year and substance_outcome. 
```{r}
animal_outcome_substance <- inner_join(animal_outcome, animal_substance_long, by = "Reference")
animal_outcome_substance <- animal_outcome_substance %>% select("Reference", "Year.x", "Substance", "species", "outcome")
colnames(animal_outcome_substance) <- c("Reference", "Year", "Substance", "Species", "Outcome")
```

Save the Excel file

```{r}
write.xlsx(animal_outcome_substance,
           file = file.path(newday,"animal_outcome_substance.xlsx"),
           colNames = TRUE, borders = "rows"
)
```


### Cleaning the human data with regard to the health outcome

```{r}
human_outcome <- human

human_outcome <- human_outcome %>% 
  select("Reference", "Year", "Outcome") # Select the coloumns of interest

```

```{r}
human_outcome <- human_outcome %>% 
  mutate(`Gastric emptying (time and frequency)`= if_else(str_detect(`Outcome`, "Gastric"), "yes", "No"))

human_outcome <- human_outcome %>% 
  mutate(`Faecal weight and consistency`= if_else(str_detect(`Outcome`, "Faecal"), "yes", "No"))

human_outcome <- human_outcome %>% 
  mutate(`Satiety and appetite`= if_else(str_detect(`Outcome`, "Satiety"), "yes", "No"))

human_outcome <- human_outcome %>% 
  mutate(`Glucose and insulin homeostasis`= if_else(str_detect(`Outcome`, "glucose"), "yes", "No"))

human_outcome <- dplyr::select(human_outcome, c(!`Outcome`))


human_outcome <- human_outcome %>% pivot_longer(
  cols = "Gastric emptying (time and frequency)":"Glucose and insulin homeostasis",
  names_to = "outcome",
  values_to = "Included"
)

human_outcome <- human_outcome %>% 
  filter(Included == "yes")

human_outcome_wide <- human_outcome %>% pivot_wider(
  names_from = "outcome",
  values_from = "Included"
)


```


Merge the animal tables with substance_year and substance_outcome. 
```{r}
human_outcome_substance <- inner_join(human_outcome, human_substance_long, by = "Reference")
human_outcome_substance <- human_outcome_substance %>% select("Reference", "Year.x", "Substance", "species", "outcome")
colnames(human_outcome_substance) <- c("Reference", "Year", "Substance", "Species", "Outcome")

```

Save the Excel file

```{r}
write.xlsx(human_outcome_substance,
           file = file.path(newday,"human_outcome_substance.xlsx"),
           colNames = TRUE, borders = "rows"
)
```


# Make figures data extraction

## Figure substance and year of publication

```{r}


counts <- animal_human_substance_year %>% 
  count(Substance, Year_cat)

marginal_counts_Year <- counts %>%
  group_by(Year_cat) %>%
  summarise(n = sum(n))

marginal_counts_Substance <- counts %>%
  group_by(Substance) %>%
  summarise(n = sum(n))

year_seq <- c("<1993", "1994-2003", "2004-2013", "2014-2024")
substance_seq <- c("Agar (E 406)", "Carboxymethyl cellulose (E 466)", "Carrageenan (E 407)", "Eucheuma seaweed (407)", "Gellan gum (E 418)", "Sodium alginate (E 401)", "Xanthan gum (E 415)")


Heat_map_subst_year <- ggplot(counts, aes(Year_cat, Substance)) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = round(n, 1))) +
  scale_fill_gradient(low = "lightblue", high = "olivedrab")+
  scale_x_discrete(limits = year_seq)+
  theme_minimal() +
  labs(x = "Time period (year)", y = "EST")+
  theme(axis.text.x = element_text(size = 11), axis.text.y = element_text(size = 11), axis.title.x = element_text(size = 16, face =  "bold"), axis.title.y = element_text(size = 16, face =  "bold"), legend.position = "none")  # Move legend below the plot

Heat_map_subst_year

p_bar_Year <- ggplot(marginal_counts_Year, aes(x = Year_cat, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab4") +
  labs(x = "Year", y = "No. of studies \nper time period across EST") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  scale_x_discrete(limits = year_seq)+
  theme_minimal() + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank(), axis.text.y = element_text(size = 11), axis.title.y = element_text(size = 16, face = "bold"))

p_bar_Year

p_bar_substance_year <- ggplot(marginal_counts_Substance, aes(x = Substance, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab") +
  labs(x = " Substance", y = "No. of studies\ntesting each EST") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  coord_flip() +
  theme_minimal() + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(size = 11), axis.title.x = element_text(size = 16, face =  "bold"))
  

p_bar_substance_year




# Combine the plots using patchwork
Heat_bar_plot <- (p_bar_Year+ plot_spacer() + plot_layout(ncol = 2, heights = c(1, 0.1))) / 
                    (Heat_map_subst_year | p_bar_substance_year + plot_spacer()) +
  plot_layout(widths = c(4, 1))

Heat_bar_plot

ggsave(filename=file.path(newday, "Heat_bar_plot.jpeg"),
       device = NULL,
       width=NA,
       height=NA,
       units="mm")




```

## Figure substance and  outcome in animals
```{r}


counts2 <- animal_outcome_substance %>% 
  count(Substance, `Outcome`)

marginal_counts_outcome <- counts2 %>%
  group_by(`Outcome`) %>%
  summarise(n = sum(n))

marginal_counts_Substance2 <- counts2 %>%
  group_by(Substance) %>%
  summarise(n = sum(n))


Heat_outcom_subs_animal <- ggplot(counts2, aes(Substance, Outcome)) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = round(n, 1))) +
  scale_fill_gradient(low = "lightblue", high = "olivedrab")+
  #scale_x_discrete(limits = year_seq)+
  theme_minimal() +
  labs(x = "EST", y = "Endpoint")+
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 0.75, hjust=0.75 ), axis.text.y = element_text(size = 12), axis.title.x = element_text(size = 16, face =  "bold"), axis.title.y = element_text(size = 14, face =  "bold"), legend.position = "none")  # Move legend below the plot

Heat_outcom_subs_animal

p_bar_outcome <- ggplot(marginal_counts_outcome, aes(x = Outcome, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab4") +
  labs(x = "Endpoint", y = "No. of studies \nper endpoint accross EST") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  coord_flip() +
  #scale_x_discrete(limits = year_seq)+
  theme_minimal() + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(size = 12), axis.title.x = element_text(size = 14, face =  "bold"))

p_bar_outcome

p_bar_Substance2 <- ggplot(marginal_counts_Substance2, aes(x = Substance, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab") +
  labs(x = "EST", y = "No. of studies per EST \naccross endpoints") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  theme_minimal() + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank(), axis.text.y = element_text(size = 12), axis.title.y = element_text(size = 14, face = "bold"))
  
  

p_bar_Substance2


# Combine the plots using patchwork
Heat_bar_outcome_animals <- (p_bar_Substance2+ plot_spacer() + plot_layout(ncol = 2, heights = c(1, 0.1))) / 
                    (Heat_outcom_subs_animal | p_bar_outcome + plot_spacer()) +
  plot_layout(widths = c(4, 1))

Heat_bar_outcome_animals

ggsave(filename=file.path(newday, "Heat_bar_outcome_animals.jpeg"),
       device = NULL,
       width=NA,
       height=NA,
       units="mm")


```

## Figure substance and  outcome in humans
```{r}


counts3 <- human_outcome_substance %>% 
  count(Substance, `Outcome`)

marginal_counts_outcome <- counts3 %>%
  group_by(`Outcome`) %>%
  summarise(n = sum(n))

marginal_counts_Substance2 <- counts3 %>%
  group_by(Substance) %>%
  summarise(n = sum(n))


Heat_outcom_subs_human <- ggplot(counts3, aes(Substance, Outcome)) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = round(n, 1))) +
  scale_fill_gradient(low = "lightblue", high = "olivedrab")+
  #scale_x_discrete(limits = year_seq)+
  theme_minimal() +
  labs(x = "EST", y = "Endpoint")+
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 0.75, hjust=0.75 ), axis.text.y = element_text(size = 12), axis.title.x = element_text(size = 14, face =  "bold"), axis.title.y = element_text(size = 14, face =  "bold"), legend.position = "none")  # Move legend below the plot

Heat_outcom_subs_human

p_bar_outcome <- ggplot(marginal_counts_outcome, aes(x = Outcome, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab4") +
  labs(x = "Endpoint", y = "No. of studies \nper endpoint accross EST") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  coord_flip() +
  #scale_x_discrete(limits = year_seq)+
  theme_minimal() + 
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(size = 12), axis.title.x = element_text(size = 14, face =  "bold"))

p_bar_outcome

p_bar_Substance2 <- ggplot(marginal_counts_Substance2, aes(x = Substance, y = n)) +
  geom_bar(stat = "identity", fill = "olivedrab") +
  labs(x = " Substance", y = "No. of studies per EST \naccross endpoint") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5))+
  theme_minimal() + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank(), axis.text.y = element_text(size = 12), axis.title.y = element_text(size = 14, face = "bold"))
  
  

p_bar_Substance2


# Combine the plots using patchwork
Heat_bar_outcome_humans <- (p_bar_Substance2+ plot_spacer() + plot_layout(ncol = 2, heights = c(1, 0.1))) / 
                    (Heat_outcom_subs_human | p_bar_outcome + plot_spacer()) +
  plot_layout(widths = c(4, 1))

Heat_bar_outcome_humans

ggsave(filename=file.path(newday, "Heat_bar_outcome_humans.jpeg"),
       device = NULL,
       width=NA,
       height=NA,
       units="mm")


```


# Make Sankey plot

```{r}


pl <- df %>% drop_na(node) %>% # This remowes the NA's before making the Sankey plot. 
  ggplot(aes(x = x,                        
                     next_x = next_x,                                     
                     node = node,
                     next_node = next_node,        
                     fill = factor(node),
                     label = node))            # This Creates a label for each node


                     
pl <- pl +geom_sankey(flow.alpha = 0.5,          #This Creates the transparency of your node 
                      node.color = "black",     # This is your node color        
                      show.legend = TRUE)        # This determines if you want your legend to show

pl <- pl + geom_sankey_label(size = 5, 
                             color = "black", 
                             fill = "white") # This specifies the Label format for each node 



pl <- pl + theme(legend.position = 'none')
pl <- pl + theme(axis.title = element_blank(),
                 axis.text.y = element_blank(),
                 axis.ticks = element_blank(),
                 panel.grid = element_blank(),
                 axis.text.x = element_text(size = 20))


pl <- pl + scale_fill_viridis_d(option = "magma")


pl


ggsave(filename=file.path(newday, "pl.jpeg"),
       device = NULL,
       width=NA,
       height=NA,
       units="mm")







```








